JD notes on effects

### 2022-07-07 manuscript/body.tex
Cape Town airport

To make an outcome plot, we use a \emph{focal} predictor on the x-axis. The resulting plot will depend on choices we make about other (non-focal) predictor(s). In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate. However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the response. Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms), which extend ordinary linear models to incorporate random effects. How we treat random effects can also have an effect.

The outcome plots described above are often called prediction plots or effects plots. We endeavor here to make a conceptual distinction. If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central estimate, which is the same for both prediction and effects plots. However, the distinction between the two lies in how we describe the uncertainties around the central estimate. In particular, we may want to capture the uncertainty in the whole model (these include uncertainty due to intercept, focal and non-focal predictors, and random effects), or we may be interested in showing the uncertainty associated with the focal predictor only. In the former case, we use the prediction plots, which generally focus on the \emph{total} effect and will often be suited for a univariate model. In the latter case, we use effect plots that focus on the direct effect of the focal predictor -- after controlling for other non-focal predictors -- from a multivariate model, which will be preferable for multivariate models. Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots. Shi et al. \citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}

Generating quantities, i.e., central estimate together with the corresponding confidence intervals has some challenges. In particular:
\begin{enumerate}
\item choice of representative values of focal predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
\item bias in the expected mean prediction induced by the non-linear transformation of the response variable (especially in GL(M)Ms).
\end{enumerate}

The most common way of choosing the representative values is taking unique levels or quantiles of the focal predictor if discrete or continuous, respectively. The central estimates are then calculated by holding the non-focal predictors at their reference point (values chosen for the non-focal predictors, e.g., means) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as the \emph{central estimates} for an ``average case'' in a population; the associated confidence intervals will either describe a prediction or effect plot. In the effect plots, we can also choose an anchor (value chosen for the focal predictor).

@@

Muli-parameter variables (MPVs) such as splines, polynomials, etc., can be within the focal predictor, or within (between) non-focal predictor(s). To distinguish the two, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
%
\begin{align}\label{eq:lm_cubic}
\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i\nonumber \\
&+ \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i.
\end{align}
%
In the first case, with \code{Age} as the focal input variable, is a cubic polynomial with three model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$). In this case, each model variable is evaluated separately across the chosen levels of the focal input variable, $Age_i$. Specifically, the model variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section. The central estimates associated with \code{Age} on the linear predictor scale become
%
\begin{align}
\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}.
\end{align}
%
In the second case, with \code{Wealthindex} as the focal predictor, the non-focal predictor, \code{Age}, is a cubic polynomial. In this case, the non-focal model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are all treated as separate non-focal model variables and an appropriate choice of reference point applies just like in the models without MPVs. For instance, in mean-based approach, we average all non-focal MPVs. Thus
%
\begin{align}
\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i.
\end{align}
%

>>>>

To make an outcome plot, we use a \emph{focal} predictor on the x-axis. The resulting plot will depend on choices we make about other (non-focal) predictor(s). In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate. However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the response. Additional challenges arise when dealing with “mixed” models, which incorporate random effects.

As noted, the outcome plots described above are often called prediction plots or effects plots. We endeavor here to make a conceptual distinction. 
The primary distinction between the two lies in how we describe the uncertainties around the central estimate. 
If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, then we may want to capture a variety of sources of uncertainty, including that due to the intercept, focal and non-focal predictors, and random effects.
Conversely, if we wish to focus on the \emph{effect} of a focal predictor only, we might want to isolate uncertainty due to coefficients associated with that predictor.
If we follow this convention, we expect effects plots to have narrower confidence intervals (CIs) than prediction plots.

The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model, that only contains terms related to that predictor.
If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects, or leave covariates out, in order to estimate “total” effects (direct plus indirect), or take an intermediate strategy.
Shi et al. \citep{shi_evidence_2017} used multivariate effects plots to visualize estimated direct effects of different predictors in a paper that compared the difference in sexual risk behaviors between circumcised and uncircumcised men.

Generating quantities, i.e., central estimate together with the corresponding confidence intervals has some challenges. In particular:
\begin{enumerate}
\item choosing the \emph{reference point} for non-focal predictors in multivariate models
\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
\item \jd{When is this a challenge? Doesn't it just work like magic with the lm machinery? and how to incorporate the uncertainty due to non-focal predictors}
\item biases induced by non-linear transformations of the response variable in generalized models (especially generalized mixed models).
\item choice of representative values of the focal predictor \jd{I still don't see this as much of a challenge.}
\end{enumerate}

Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor, or an exhaustive set of levels for a categorical predictor.
The central estimates are then calculated by holding the non-focal predictors at their reference point (values chosen for the non-focal predictors, e.g., means) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as the \emph{central estimates} for an ``average case'' in a population; the associated confidence intervals will either describe a prediction or effect plot. In the effect plots, we can also choose an anchor (value chosen for the focal predictor).

@@

Multi-parameter variables (MPVs) such as splines, polynomials, etc., can be within the focal predictor, or within non-focal predictor(s). To distinguish the two, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
%
\begin{align}\label{eq:lm_cubic}
\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i\nonumber \\
&+ \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i.
\end{align}
%
In the first case, with \code{Age} as the focal input variable, is a cubic polynomial with three model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$). In this case, each model variable is evaluated separately across the chosen levels of the focal input variable, $Age_i$. Specifically, the model variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section. The central estimates associated with \code{Age} on the linear predictor scale become
%
\begin{align}
\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}.
\end{align}
%
In the second case, with \code{Wealthindex} as the focal predictor, the non-focal predictor, \code{Age}, is a cubic polynomial. In this case, the non-focal model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are all treated as separate non-focal model variables and an appropriate choice of reference point applies just like in the models without MPVs. For instance, in mean-based approach, we average all non-focal MPVs. Thus
%
\begin{align}
\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i.
\end{align}
%
